---
title: "Day 7"
output: github_document
---
## Part 1

The goal is to find the name of the program at the bottom of the pile. The input looks like below. This is an example, we are supposed to find `tknk` as the answer.


```{r message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(stringr)
library(testthat)

example_input <-  'pbga (66)
xhth (57)
ebii (61)
havc (66)
ktlj (57)
fwft (72) -> ktlj, cntj, xhth
qoyq (66)
padx (45) -> pbga, havc, qoyq
tknk (41) -> ugml, padx, fwft
jptl (61)
ugml (68) -> gyxo, ebii, jptl
gyxo (61)
cntj (57)'
```

Let's pre process the example a bit, so as we have a nice character vector with an entry for each line.

```{r}
example_input <- example_input %>% 
  str_split('\\n') %>% 
  pluck(1)
example_input
```

Now, I will find all the program names and their weight.

```{r}
 get_name_weight <- function(input){
   input %>% 
     str_match('([a-z]+) \\((\\d+)\\)') %>% 
     as_tibble() %>% 
     select("name" = V2, "weight" = V3) %>% 
     mutate(weight = as.numeric(weight))
}
prog_names <- get_name_weight(example_input)
prog_names
```

Ok now, let's find the programs that are in towers, i.e. the programs whose names are after the `"->"` in the input. I flatten the result, so as to have a single character vector.

```{r}
get_towers <- function(input){
  input %>% 
    str_subset('->') %>%    # Get only the lines where there is a '->'
    str_split('-> ') %>%    # Split the line
    map(2) %>%              # Get the values after the '-> '
    str_split(', ') %>%     # Split the lines after commas
    flatten_chr()           # Flatten the result
}
prog_in_towers <- get_towers(example_input)
prog_in_towers
```

Now, the bottom of the tower is the name that is not in the above list.

```{r}
prog_names %>% 
  filter(!(name %in% prog_in_towers)) %>% 
  `[[`('name')
```

That works! Let's make it a function for use with the real file.
```{r}
get_bottom <- function(input){
  input %>% 
    get_name_weight() %>% 
    filter(!(name %in% get_towers(input))) %>% 
    `[[`('name')
}
expect_equal(get_bottom(example_input), "tknk")
```

Now with the file
```{r}
my_input <- read_lines("day07-input.txt")

get_bottom(my_input)
```

## Part 2
Now we have to find which tower is not balanced. The weights of all the towers above one disk must be equal. I want to have the children (if any), and I split it to have a list with the names. `children_list` is a list and it is a little akward to work with afterwards, but I couldn't do better...

```{r}
get_name_weight_children <- function(input){
   input %>% 
     str_match('([a-z]+) \\((\\d+)\\)(?: -> )?(.*)?') %>% 
     as_tibble() %>% 
     select("name" = V2, "weight" = V3, "children" = V4) %>% 
     mutate(weight = as.numeric(weight)) %>% 
     mutate(children_list = str_split(children, ', '))
 }

prog_weight <- get_name_weight_children(example_input)
prog_weight
```

I then want to have the total weight for a given program name. `get_total_weight` is a recursive function that gives either the weight if the program do not have children, or the sum of his own weight and the weight of all the children.

```{r}
get_total_weight <- function(name){
  i <- which(prog_weight$name == name) # Thanks to jennybc for the trick to index the list using an index
  weight = prog_weight$weight[i]
  if(prog_weight$children[i] == ""){
    return(weight)
  }
  else {
    return(weight + sum(map_dbl(prog_weight$children_list[[i]], 
                                get_total_weight)))
  }
}

expect_equal(get_total_weight('xhth'), 57)
expect_equal(get_total_weight('fwft'), 243)
expect_equal(get_total_weight('ugml'), 251)
```

It works on the example, nice!
Next, I want to check wether the program is balanced, i.e. if all the children have the same total weight. If the program doesn't have any children, then it is `TRUE`, if not, I compute the weights of all the children in the list, and check if the min and the max are the same (I compute `diff(range()) == 0` which is equivalent).

```{r}
balanced_above <- function(name) {
  i <- which(prog_weight$name == name)
  if(prog_weight$children[i] == ""){
    return(TRUE)
  }
  else {
    above_weights = map_dbl(prog_weight$children_list[[i]],
                            get_total_weight)
    return(diff(range(above_weights)) == 0)
  }
}

expect_true(balanced_above('xhth'))
expect_true(balanced_above('fwft'))
expect_false(balanced_above('tknk'))
```

To find which program is unbalanced, we have to move up from the root. If all the children are balanced, we're done, if not, we find the one that is different and check again.

```{r}
root <- get_bottom(example_input)

find_topmost_unbalanced <- function(root){
    children_names <- prog_weight %>% 
      filter(name == root) %>%
      `$`(children_list) %>%
      pluck(1)
    
    child_balance <- map_lgl(children_names, balanced_above)
    
    if(all(child_balance)){
      return(root)
    }
    else {
      return(find_topmost_unbalanced(children_names[! child_balance]))
    }
}

expect_equal(find_topmost_unbalanced(root), "tknk")
```

The next function gives a tibble with the weights of the children above, both total and individual.

```{r}
weights_above <- function(name) {
  i <-  which(prog_weight$name == name)
  weights <- map_dbl(prog_weight$children_list[[i]], get_total_weight)
  tibble(name = prog_weight$children_list[[i]], 
         total_weight = weights) %>% 
    inner_join(prog_weight %>% select(name, weight), by = "name")
}
weights_above('tknk')
```

Given those weights for the topmost unbalanced program, we compute the difference between the total weights and the median weight. This gives by how much the program is unbalanced. The difference between the individual weight and this offset gives the weight the program should have in order to balance everything. For the example, we get `60`, which is the expected result.

```{r}
weights_above(find_topmost_unbalanced(root)) %>% 
  mutate(diff = total_weight - median(total_weight),
         balance = weight - diff) %>% 
  filter(diff != 0)
```

To compute if for the real input, I have the compute `prog_weight`, find the root and apply the same pipe as above.

```{r}
prog_weight <- get_name_weight_children(my_input)
root <- get_bottom(my_input)

weights_above(find_topmost_unbalanced(root)) %>% 
  mutate(diff = total_weight - median(total_weight),
         balance = weight - diff) %>% 
  filter(total_weight != median(total_weight))
```

The answer is `1674`, which is correct!